Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-05 ms/manuscript.Rmd
Update MS to use mpd, mntd instead of FRic etc

---
title: "Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
---

@@

We analyzed community phylogenetic diversity using the Mean Phylogenetic Distance (MPD) and Mean Nearest Taxon Distance (MNTD) indices of [@Webb2000]. MPD is the mean phylogenetic distance of each species in a community to all other species in the same community; MNTD is the mean phylogenetic distance of each species in a community to its sister species in the same community. Communities with low MPD or MNTD are 'phylogenetically clustered' (co-occurring species are closely related to each other), whereas those with high MPD or MNTD are 'phylogenetically overdispersed' (co-occurring species are distantly related to each other). MPD therefore provides a measure of phylogenetic diversity across the whole tree, while MNTD is more sensitive to distances towards the tips. Alone, MPD and MNTD are not very informative; to assess their statistical significance, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{1000} randomized communities (equivalent to the Net Relatedness Index and Nearest Taxon Index, respectively, of @Webb2000 with the sign reversed). Observed values above 95\% or less than 5\% of the null distribution were considered significant. Selection of the randomization procedure can strongly bias outcomes [@Gotelli2006;@Ulrich2012]. Our null hypothesis is that fern spores are easily capable of dispersal between terrestrial and epiphytic habitats, and across the relatively short distances separating our plots. We therefore constructed null communities by drawing species randomly from a regional pool including terrestrial and epiphytic taxa. The regional pool included all known species of ferns from Moorea, and all species from the neighboring island of Tahiti with genetic data available ($\n = 146$ \si{\species} total; [@Nitta2017]). We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate community phylogenetic diversity metrics.

We analyzed community functional trait diversity using the \function{dbFD} function in the \package{FD} R package [@Laliberte2011]. This function takes as input values any number of traits for species in a given community, plots them in a multidimensional trait space using PCA, and calculates several summary statistics. Functional richness (FRic) is equivalent to convex hull volume containing all observed trait values plotted in multidimensional trait space [@Cornwell2006], functional evenness (FEve) is the average distance to the center of all the traits, and functional dispersion (FDis) is the mean distance between traits [@Villeger2008]. We chose these three metrics because they are statistically independent and can each provide different information about the structure of community trait diversity [@Savage2012]. Functional richness does not use abundance data, but functional evenness and divergence can be weighted by abundance, so we preformed analyses for these metrics with abundance weighting to correct for the effect of rare species with unusual morphologies. Selection, transformation, and scaling of traits was performed similarly as for the PCA, except that the method used to calculate distance [@Gower1971] can accommodate missing data, so we included species with observations missing for some traits.

We checked for correlation between diveristy metrics with Pearson's correlation coefficient, and retained a set of metrics with absolute correlation coefficients < 0.9. We tested for differences in diversity metrics between epiphytic and terrestrial communities with a paired \tval-test. We simultaneously tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_select) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

>>>>

---
title: "Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
---

@@

We used distance-based methods to analyze phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000]. These metrics were originally formulated for phylogenetic distances (thus "phylogenetic" in the name), but can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. For functional diversity, we used Gower's dissimilarity coeffecient [@Gower1971] because it can use continuous, binary, and qualitative traits, and accepts missing data. We log-transformed traits as needed to account for skew, excluded frond width and length (both frond size traits highly correlated with stipe length), and weighted sporophyte and gametophyte traits equally before calculating the distance matrix using the \function{gowdis} function in the R pacakge \package{FD}.

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities (equivalent to the Net Relatedness Index and Nearest Taxon Index, respectively, of @Webb2000 with the sign reversed). Selection of the randomization procedure can strongly bias outcomes [@Gotelli2006;@Ulrich2012]. Our null hypothesis is that fern spores are easily capable of dispersal between terrestrial and epiphytic habitats, and across the relatively short distances separating our plots. We therefore constructed null communities using the "independent swap" algorithm of @Gotelli2000 that maintains species frequency and abundance while randomizing species identities. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics.

We tested for differences in diversity metrics between epiphytic and terrestrial communities with a paired \tval-test. We simultaneously tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_select) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].
